---
title: "Mortgage Lending Decisions"
author: "Mark Russeff, Renato Albolea, Shuai Ma"
subtitle: Summary Findings
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---
&nbsp;
&nbsp;


\centerline{Controlling for relevant characteristics, is race/ethnicity associated with the outcome of a mortgage loan application?} 
  
&nbsp;
&nbsp;


\begin{table}[!htbp] \centering 
  \caption{Descriptive statistics - \textbf{All}} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Approved & 0.88 & 0.33 & 0 & 1 \\ 
Meets credit guidelines & 0.91 & 0.28 & 0 & 1 \\ 
Other obligations (\% of income) & 32.37 & 8.25 & 0 & 95 \\ 
Loan to purchase price (\%) & 76.08 & 16.76 & 2 & 100 \\ 
Married & 0.66 & 0.47 & 0 & 1 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}
```{r}

```

\begin{table}[!htbp] \centering 
  \caption{Descriptive statistics - \textbf{Race}} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
\textbf{White} (n = 1641)\\
\hspace{2ex}Approved & 0.91 & 0.29 & 0 & 1 \\ 
\hspace{2ex}Meets credit guidelines & 0.94 & 0.24 & 0 & 1 \\ 
\hspace{2ex}Other obligations (\% of income) & 31.99 & 8.18 & 0 & 95 \\ 
\hspace{2ex}Loan to purchase price (\%) & 74.78 & 17.17 & 2 & 100 \\ 
\hspace{2ex}Married & 0.66 & 0.47 & 0 & 1 \\ 
\textbf{Black} (n = 192)\\
\hspace{2ex}Approved & 0.67 & 0.47 & 0 & 1 \\ 
\hspace{2ex}Meets credit guidelines & 0.72 & 0.45 & 0 & 1 \\ 
\hspace{2ex}Other obligations (\% of income) & 35.03 & 8.13 & 5.60 & 63 \\ 
\hspace{2ex}Loan to purchase price (\%) & 82.89 & 12.63 & 28.99 & 100 \\ 
\hspace{2ex}Married & 0.61 & 0.49 & 0 & 1 \\ 
\textbf{Hispanic} (n = 104)\\
\hspace{2ex}Approved & 0.78 & 0.42 & 0 & 1 \\ 
\hspace{2ex}Meets credit guidelines & 0.87 & 0.34 & 0 & 1 \\ 
\hspace{2ex}Other obligations (\% of income) & 33.32 & 8.56 & 14.60 & 62 \\ 
\hspace{2ex}Loan to purchase price (\%) & 83.91 & 10.92 & 40 & 100 \\ 
\hspace{2ex}Married & 0.71 & 0.46 & 0 & 1 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

```{r}

```
\begin{table}[!htbp] \centering 
  \caption{Logit Model} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{2}{c}{\textit{Dependent variable:}} \\ 
\cline{2-3} 
\\[-1.8ex] & \multicolumn{2}{c}{Approve} \\ 
\\[-1.8ex] & Estimate & Odds Ratio\\ 
\hline \\[-1.8ex] 
 Meets credit guidelines & 3.766$^{***}$ & 43.227 \\ 
  & (0.221) \\ 
  & & \\ 
 Other obligations (\% of income) & $-$0.034$^{***}$ & 0.967 \\ 
  & (0.011) \\ 
  & & \\ 
 Loan to purchase price (\%) & $-$0.016$^{**}$ & 0.984 \\ 
  & (0.007) \\ 
  & & \\ 
 Married & 0.482$^{***}$ & 1.619 \\ 
  & (0.185) \\ 
  & & \\ 
 Black & $-$0.869$^{***}$ & 0.419 \\ 
  & (0.243) \\ 
  & & \\ 
 Hispan & $-$0.860$^{***}$ & 0.423 \\ 
  & (0.323) \\ 
  & & \\ 
 Constant & 1.233$^{*}$ & 3.431 \\ 
  & (0.685) \\ 
  & & \\ 
\hline \\[-1.8ex] 
Observations & 1,937 \\ 
Log Likelihood & $-$462.546 \\ 
Akaike Inf. Crit. & 939.091 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\textit{Reference category:} white\\
Standar Errors in parenthesis\\ 
\end{tabular} 
\end{table} 

```{r}

```
\begin{table}[!htbp] \centering 
  \caption{Predicted Probability of Loan Approval - \textbf{Logit}} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
& \multicolumn{1}{c}{White} & \multicolumn{1}{c}{Black} & \multicolumn{1}{c}{Hispanic} \\ 
\hline \\[-1.8ex] 
Meets credit guidelines, Married & 96.0\% & 90.9\% & 91.0\% \\
Meets credit guidelines, Single & 93.7\% & 86.1\% & 86.2\% \\ 
Does not meet credit guidelines, Married & 35.6\% & 18.8\% & 19.0\% \\ 
Does not meet credit guidelines, Single & 25.5\% & 12.5\% & 12.6\% \\ 
\hline \\[-1.8ex] 
\textit{Notes:} Other obligations set at sample mean of 32.37\% \\
Loan to purchase price set at sample mean of 76.08\% \\
\end{tabular} 
\end{table}
```{r}

```

\begin{table}[!htbp] \centering 
  \caption{Estimated Probit Model} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Race & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
\cline{2-2} 
\\[-1.8ex] & Approve \\ 
\hline \\[-1.8ex] 
  Meets credit guidelines & 2.169$^{***}$ \\ 
  & (0.123) \\ 
  & \\
 Other obligations (\% of income) & $-$0.016$^{***}$ \\ 
  & (0.005) \\ 
  & \\ 
 Loan to purchase price (\%) & $-$0.007$^{**}$ \\ 
  & (0.003) \\ 
  & \\ 
 Married & 0.239$^{***}$ \\ 
  & (0.092) \\ 
  & \\ 
 Black & $-$0.450$^{***}$ \\ 
  & (0.128) \\ 
  & \\ 
 Hispanic & $-$0.444$^{***}$ \\ 
  & (0.169) \\ 
  & \\ 
 Constant & 0.435 \\ 
  & (0.337) \\ 
  & \\ 
\hline \\[-1.8ex] 
Observations & 1,937 \\ 
Log Likelihood & $-$462.491 \\ 
Akaike Inf. Crit. & 938.982 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\textit{Reference category:} white\\
Standar Errors in parenthesis\\
\end{tabular} 
\end{table} 
```{r}

```
\begin{table}[!htbp] \centering 
  \caption{Predicted Probability of Loan Approval - \textbf{Probit}} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
& \multicolumn{1}{c}{White} & \multicolumn{1}{c}{Black} & \multicolumn{1}{c}{Hispanic} \\ 
\hline \\[-1.8ex] 
Meets credit guidelines, Married & 96.1\% & 90.4\% & 90.5\% \\
Meets credit guidelines, Single & 93.6\% & 85.7\% & 85.9\% \\ 
Does not meet credit guidelines, Married & 34.0\% & 19.4\% & 19.6\% \\ 
Does not meet credit guidelines, Single & 25.8\% & 13.5\% & 13.7\% \\ 
\hline \\[-1.8ex] 
\textit{Notes:} Other obligations set at sample mean of 32.37\% \\
Loan to purchase price set at sample mean of 76.08\% \\
\end{tabular} 
\end{table}
