---
title: "MLA"
author: "Mark Russeff, Renato Albolea, Shuai Ma"
subtitle: "Primary Findings"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

# Load Packages
library(here)
library(tidyverse)
library(janitor)
library(aod)
library(htmlTable)
```

# 1. Abstract

TBD

# 2. Background

The following abstract appeared in Alicia H. Munnell, Geoffrey M.B. Tootell, Lynn E. Browne, and James McEneaney (1996), “Mortgage Lending in Boston: Interpreting HMDA Data,” American Economic Review 86, 25-53.

>_The Home Mortgage Disclosure Act was enacted to monitor minority and low-income access to the mortgage market. The data collected for this purpose show that minorities are more than twice as likely to be denied a mortgage as whites.  Yet variables correlated with both race and creditworthiness were omitted from these data, making any conclusion about race’s role in mortgage lending impossible. The Federal Reserve Board of Boston collected_ additional variables important to the mortgage lending decision….

As discussed in Munnell et al (1996), the HMDA data indicate whether an applicant’s mortgage application was approved and provide several demographic characteristics.  In 1990, following the request of the Federal Reserve Board of Boston, lending institutions in the Boston area provided additional information relevant to mortgage lending decisions.  In light of the relatively small number of mortgage loan applications made by minorities, these extra variables were collected for all applications by blacks and Hispanics and for a random sample of those by whites.

>_All applicants are non-Hispanic white, non Hispanic black, or Hispanic.  [In 1990 about 94% of Boston residents were white, Black, or Hispanic.](http://www.bostonplans.org/getattachment/83972a7a-c454-4aac-b3eb-02e1fddd71e3/)_

# 3. Research Question

  * __Controlling for relevant characteristics, is race/ethnicity associated with the outcome of a mortgage loan application?__

# 4. Explortory Data Analysis

## 4.1 Data Wrangling

```{r load_data, include=FALSE}
base <- read_csv(here('raw_data','MLD Data File.csv'))
```

__(A) First View of Data__ (The code and result are hid here since this is not key point of this paper, and makes this paper clean. The final data summary will be showed.)
```{r include=FALSE}
str(base)
summary(base)
```

1. First Thought: 

  * The following column names should be updated:
    + GDLIN - Credit_History (credit history meets the guideline)
    + OBRAT - Other_Obligations (other obligations as a percent of total income) 
    + LOANPRC - Loan_Percentage (loan amount/purchase price)
    
  * The following variables should be changed to factors:
    + MARRIED
    + GDLIN
    + BLACK
    + HISPAN
    + MALE
    + APPROVE

```{r include=FALSE}
names(base) <- c("Married", "Credit_History" , "Other_Obligations" , "Black" , "Hispan" , "Male" , "Approve" , "Loan_Percentage")
base$Married <- as.factor(base$Married)
base$Credit_History <- as.factor(base$Credit_History)
base$Black <- as.factor(base$Black)
base$Hispan <- as.factor(base$Hispan)
base$Male <- as.factor(base$Male)
base$Approve <- as.factor((base$Approve))

#Change variable Loan_Percentage to a percentage
base <- base %>%
  mutate(Loan_Percentage = Loan_Percentage * 100)
```

__(B) Second View of Data__ (Code and results are hid.)
```{r include=FALSE}
str(base)
summary(base)
```

2. Second Thought (criteria for sub-setting data ):

  * Remove the three "." from "Married"
  * Remove the two "666" from "Credit_History"
  * Remove the fifteen "." from "Male"
  * Remove the value which higher than 1 from "Loan_Percentage"

```{r include=FALSE}
base <- subset(base, Loan_Percentage <= 100 & Male != "." & Married != "." & Credit_History != 666)
base <- droplevels(base)
```

__(C) Final Data Summary__
```{r}
summary(base)
```



## 4.2 Race and Gender 

(Code is hid from this paper, but available based on request)

__Findings from race and gender: __

  1. The proportion of Black and Hispanic applications account for 10% and 5% of all applications separately. 
  2. The proportion of Male and Female applications account for 19% and 81% of all applications separately. 
  3. The smallest category in the application is "Hispanic Female" which was 1%. 
  4. The largest category in the application is "Non-Black/Hispanic Male" which was 70%. 

```{r include=FALSE}
rg <- base %>% 
  select(Hispan,Black,Male)

rg$Hispan <- as.character(rg$Hispan)
rg$Hispan[rg$Hispan == "1"] <- "Hispan"
rg$Hispan[rg$Hispan == "0"] <- NA

rg$Black <- as.character(rg$Black)
rg$Black[rg$Black == "1"] <- "Black"
rg$Black[rg$Black == "0"] <- NA

rg$Male <- as.character(rg$Male)
rg$Male[rg$Male == "1"] <- "Male"
rg$Male[rg$Male == "0"] <- "Female" 

rg$Race = ifelse(is.na(rg$Hispan), rg$Black , rg$Hispan) 
rg$Race = ifelse(is.na(rg$Race), "Not Hispan/Black" , rg$Race) 
colnames(rg)[colnames(rg)=="Male"] <- "Gender"
rg$Hispan <- NULL
rg$Black <- NULL
rg$Gender <- as.factor(rg$Gender)
rg$Race <- as.factor(rg$Race)
```


```{r warning=FALSE, echo=FALSE}
rg_table_n <- rg %>% 
  tabyl(Gender,Race) %>% 
  adorn_totals(where = c("row", "col"))

htmlTable(rg_table_n,
          caption = "Total Number by Race and Gender",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2"))
```

```{r warning=FALSE, echo=FALSE}
rg_table_p <- rg %>% 
  tabyl(Gender,Race) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  adorn_percentages(denominator = "all") %>%  
  adorn_pct_formatting(digits = 1)
   
htmlTable(rg_table_p, 
          caption = "Percentage by Race and Gender",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2"))
```

## 4.3 Race and Approval

(Code is hid from this paper, but available based on request)

__Findings from race and decision: __

  1. Among all applicants the approved application from Black counts for 6.6%. The approval rate for applications from Black was __66.7%__.
  2. Among all applicants the approved application from Hispanic counts for 4.2%. The approval rate for applications from Hispanic was __77.9%__.
  3. Among all applicants the approved application from Non-Black/Hispanic counts for 77.1%.The approval rate for applications from Non-Black\Hispanic was __91.0%__.
  4. The average approval rate was __87.9%__.

```{r include=FALSE}
ra <- base %>% 
  select(Hispan,Black,Approve)

ra$Hispan <- as.character(ra$Hispan)
ra$Hispan[ra$Hispan == "1"] <- "Hispan"
ra$Hispan[ra$Hispan == "0"] <- NA

ra$Black <- as.character(ra$Black)
ra$Black[ra$Black == "1"] <- "Black"
ra$Black[ra$Black == "0"] <- NA

ra$Approve <- as.character(ra$Approve)
ra$Approve[ra$Approve == "1"] <- "Approved"
ra$Approve[ra$Approve == "0"] <- "Denied" 

ra$Race = ifelse(is.na(ra$Hispan), ra$Black , ra$Hispan) 
ra$Race = ifelse(is.na(ra$Race), "Not Hispan/Black" , ra$Race) 
#colnames(ra)[colnames(ra)=="Approve"] <- "Approve"
ra$Hispan <- NULL
ra$Black <- NULL
ra$Approve <- as.factor(ra$Approve)
ra$Race <- as.factor(ra$Race)
```


```{r warning=FALSE, echo=FALSE}
ra_table_n <- ra %>% 
  tabyl(Approve, Race) %>% 
  adorn_totals(where = c("row", "col"))

htmlTable(ra_table_n,
          caption = "Total Number by Race and Decision",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2"))
```


```{r warning=FALSE, echo=FALSE}
ra_table_p <- ra %>% 
  tabyl(Approve, Race) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  adorn_percentages(denominator = "all") %>%  
  adorn_pct_formatting(digits = 1)
   
htmlTable(ra_table_p, 
          caption = "Percentage by Race and Decision",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2"))
```

## 4.4 Race and Credit History

(Code is hid from this paper, but available based on request)

__Findings from race and credit history: __

  1. Among all applicants that the Black applicant with good credit history counts for __7.2%__. 72.3% Black applicants had good credit history.
  2. Among all applicants that the Hispanic applicants with good credit history counts for __4.6%__. 86.5% Hispanic applicants had good credit history.
  3. Among all applicants that the Non-Black/Hispanic applicants with good credit history counts for __79.6%__. 93.9% Non-Black/Hispanic applicants had good credit history.
  4. The average approval rate was __91.4%__.
  
  
```{r include=FALSE}
rc <- base %>% 
  select(Hispan,Black,Credit_History)

rc$Hispan <- as.character(rc$Hispan)
rc$Hispan[rc$Hispan == "1"] <- "Hispan"
rc$Hispan[rc$Hispan == "0"] <- NA

rc$Black <- as.character(rc$Black)
rc$Black[rc$Black == "1"] <- "Black"
rc$Black[rc$Black == "0"] <- NA

rc$Credit_History <- as.character(rc$Credit_History)
rc$Credit_History[rc$Credit_History == "1"] <- "Good"
rc$Credit_History[rc$Credit_History == "0"] <- "Not Good" 

rc$Race = ifelse(is.na(rc$Hispan), rc$Black , rc$Hispan) 
rc$Race = ifelse(is.na(rc$Race), "Not Hispan/Black" , rc$Race) 
#colnames(ra)[colnames(ra)=="Approve"] <- "Approve"
rc$Hispan <- NULL
rc$Black <- NULL
rc$Credit_History <- as.factor(rc$Credit_History)
rc$Race <- as.factor(rc$Race)
```


```{r warning=FALSE, echo=FALSE}
rc_table_n <- rc %>% 
  tabyl(Credit_History, Race) %>% 
  adorn_totals(where = c("row", "col"))

htmlTable(rc_table_n,
          caption = "Total Number by Race and Credit History",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2"))
```


```{r warning=FALSE, echo=FALSE}
rc_table_p <- rc %>% 
  tabyl(Credit_History, Race) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  adorn_percentages(denominator = "all") %>%  
  adorn_pct_formatting(digits = 1)
   
htmlTable(rc_table_p, 
          caption = "Percentage by Race and Credit History",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2"))
```

## 4.5 Race, Credit History, Decision

```{r include=FALSE}
rcd <- base %>% 
    select(Hispan,Black,Credit_History, Approve)
rcd$Hispan <- as.character(rcd$Hispan)
rcd$Hispan[rcd$Hispan == "1"] <- "Hispan"
rcd$Hispan[rcd$Hispan == "0"] <- NA

rcd$Black <- as.character(rcd$Black)
rcd$Black[rcd$Black == "1"] <- "Black"
rcd$Black[rcd$Black == "0"] <- NA

rcd$Credit_History <- as.character(rcd$Credit_History)
rcd$Credit_History[rcd$Credit_History == "1"] <- "Good"
rcd$Credit_History[rcd$Credit_History == "0"] <- "Not Good" 

rcd$Approve <- as.character(rcd$Approve)
rcd$Approve[rcd$Approve == "1"] <- "Approved"
rcd$Approve[rcd$Approve == "0"] <- "Denied" 

rcd$Race = ifelse(is.na(rcd$Hispan), rcd$Black , rcd$Hispan)
rcd$Race = ifelse(is.na(rcd$Race), "Not Hispan/Black" , rcd$Race)

rcd$Hispan <- NULL
rcd$Black <- NULL

```

```{r include=FALSE}
rcd_n_1 <- rcd %>% 
  group_by(Race, Credit_History, Approve) %>% 
  summarise(count = n())


rcd_n_2 <- rcd %>% 
  group_by(Race, Credit_History) %>% 
  summarise(count = n())


rcd_n <- left_join(rcd_n_1,rcd_n_2, by = c("Race" = "Race", "Credit_History" = "Credit_History"))
rcd_n$Category <- paste(rcd_n$Race, "-", rcd_n$Credit_History)
rcd_n$Rate <- rcd_n$count.x/rcd_n$count.y
rm(rcd_n_1)
rm(rcd_n_2)

colnames(rcd_n)[colnames(rcd_n)=="count.x"] <- "Counts"
colnames(rcd_n)[colnames(rcd_n)=="Approve"] <- "Decision"

rcd_n$count.y <- NULL
rcd_n$Race <- NULL
rcd_n$Credit_History <- NULL

rcd_n <- rcd_n[c("Category", "Decision", "Counts" , "Rate")]
rcd_n$Rate <- format(round(rcd_n$Rate, 3), nsmall = 3)
```

__Findings from race, credit history, and decision: __

  1. Among all applicants who were with "Good" credit history, the approval rate was __"Not Hispan/Black" > "Black" > "Hispan"__.
  2. Among all applicants who were with "Not Good" credit history, the approval rate was __"Not Hispan/Black" > "Hispan" > "Black"__.


__(A)Approveal/Denial Rate with Good Credit History by Races__
```{r echo=FALSE}
htmlTable(filter(rcd_n, Category == "Black - Good"),
          caption = "Approval Rate by Race and Credit History - Black with Good Credit",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          # col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2")
)
```

```{r echo=FALSE}
htmlTable(filter(rcd_n, Category == "Hispan - Good"),
          caption = "Approval Rate by Race and Credit History - Hispan with Good Credit",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          # col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2")
)
```


```{r echo=FALSE}
htmlTable(filter(rcd_n, Category == "Not Hispan/Black - Good"),
          caption = "Approval Rate by Race and Credit History - Hispan with Good Credit",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          # col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2")
)
```


__(B)Approveal/Denial Rate with Not Good Credit History by Races__
```{r echo=FALSE}
htmlTable(filter(rcd_n, Category == "Black - Not Good"),
          caption = "Approval Rate by Race and Credit History - Black with Not Good Credit",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          # col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2")
)
```

```{r echo=FALSE}
htmlTable(filter(rcd_n, Category == "Hispan - Not Good"),
          caption = "Approval Rate by Race and Credit History - Black with Not Good Credit",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          # col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2")
)
```

```{r echo=FALSE}
htmlTable(filter(rcd_n, Category == "Not Hispan/Black - Not Good"),
          caption = "Approval Rate by Race and Credit History - Black with Not Good Credit",
          #header = paste(c("Manufacture" , "Big" , "Medium/Small" , "No Ad" , "Total")),
          # col.columns = c("none", "#F1F0FA"),
          col.rgroup = c("none", "azure2")
)
```

## 4.6 Distribution of loan rate 

* loan rate = loan amount / purchase price

__Findings from distribution of loan rate by races and genders: __

  1. Fig 1 and Fig 2 below indicate that, regardless the race and gender, most of applicants applies for the loan amount around 75% to 90% of the purchase price.

```{r echo=FALSE, fig.width=8}
lp <- base %>% 
  select(Black, Hispan, Loan_Percentage)

lp$Hispan <- as.character(lp$Hispan)
lp$Hispan[lp$Hispan == "1"] <- "Hispan"
lp$Hispan[lp$Hispan == "0"] <- NA

lp$Black <- as.character(lp$Black)
lp$Black[lp$Black == "1"] <- "Black"
lp$Black[lp$Black == "0"] <- NA

lp$Race = ifelse(is.na(lp$Hispan), lp$Black , lp$Hispan)
lp$Race = ifelse(is.na(lp$Race), "Not Hispan/Black" , lp$Race)

lp$Hispan <- NULL
lp$Black <- NULL


ggplot(data = lp, aes(x = Loan_Percentage)) + geom_histogram(bins = 45, fill="deepskyblue2") + theme_bw() + facet_grid(Race ~ .) + labs(title = "Fig 1. Distribution of Loan Rate by Races", subtitle = "Loan Rate = Loan Amount / Purchase Price", x = "Loan Rate" , y = "Counts")
```


```{r echo=FALSE, fig.width=8, warning=FALSE}
lp_g <- base %>% 
  select(Male, Loan_Percentage)


lp_g$Male <- as.character(lp_g$Male)
lp_g$Gender[lp_g$Male == "1"] <- "Male"
lp_g$Gender[lp_g$Male == "0"] <- "Female"
lp_g$Gender <- as.factor(lp_g$Gender)

lp_g$Male <- NULL



ggplot(data = lp_g, aes(x = Loan_Percentage)) + geom_histogram(bins = 45, fill="deepskyblue2") + theme_bw() + facet_grid(Gender ~ .) + labs(title = "Fig 2. Distribution of Loan Rate by Gender", subtitle = "Loan Rate = Loan Amount / Purchase Price", x = "Loan Rate" , y = "Counts")
```

# 5. Estimating Models

## 5.1 Model considering Race

__(A) Estimate Logit Model__
```{r warning=FALSE}
LogitModel = glm(Approve ~ Other_Obligations + Credit_History + Loan_Percentage + Married + Black + Hispan, data = base , family = "binomial")

summary(LogitModel)
```

__The signals of the model suggest that:__

  1.	Increases in Other Obligation diminish the likelihood of approval of the loan, which makes sense.  
  2.	Being Black or Hispanic diminish your likelihood of approval of the loan, which does not make sense (your ethnicity does not interfere to your capacity to pay a loan), but was expected.  

__(B) Generate Odds Ratios__
```{r warning=FALSE}
cbind(exp(coef(LogitModel)), 1)
```

__The Odd ratios suggest that:__

1.	White people had `r round(1/exp(coef(LogitModel))[3],digits=2)` times the odds of being approved for a loan as a black person;(p<0.01)   
2.	White people had `r round(1/exp(coef(LogitModel))[4],digits=2)` times the odds of being approved for a loan as a Hispanic person;(p<0.01)  
3.	One unit of increase in other obligations reduce in `r round((1-exp(coef(LogitModel))[2])*100,digits=2)` percentage the odds of being approved for a loan in ;(p<0.01)    

__(C) Define prototypical loan applicants (you will need more than 3)__
```{r warning=FALSE}

prototype_1 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "1", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 1, Black = 0, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_2 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "1", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 1, Black = 0, Hispan = 1) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_3 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "1", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 1, Black = 1, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_4 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "1", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 0, Black = 0, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_5 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "1", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 0, Black = 0, Hispan = 1) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_6 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "1", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 0, Black = 1, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_7 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "0", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 0, Black = 0, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_8 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "0", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 1, Black = 0, Hispan = 1) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_9 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "0", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 1, Black = 1, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_10 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "0", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 1, Black = 0, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_11 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "0", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 0, Black = 0, Hispan = 1) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

prototype_12 <- data.frame(Other_Obligations = mean(base$Other_Obligations), Married = "0", Loan_Percentage = mean(base$Loan_Percentage), Credit_History = 0, Black = 1, Hispan = 0) %>%
  mutate(Black = as.factor(Black), Hispan = as.factor(Hispan), Credit_History = as.factor(Credit_History))

```

__(D) Predict probabilities for prototypical individuals__
```{r warning=FALSE}

prototype_1$predictedprob <- round(predict(LogitModel, newdata = prototype_1, type = "response")*100, digits = 1)

prototype_2$predictedprob <- round(predict(LogitModel, newdata = prototype_2, type = "response")*100, digits = 1)

prototype_3$predictedprob <- round(predict(LogitModel, newdata = prototype_3, type = "response")*100, digits = 1)

prototype_4$predictedprob <- round(predict(LogitModel, newdata = prototype_4, type = "response")*100, digits = 1)

prototype_5$predictedprob <- round(predict(LogitModel, newdata = prototype_5, type = "response")*100, digits = 1)

prototype_6$predictedprob <- round(predict(LogitModel, newdata = prototype_6, type = "response")*100, digits = 1)

prototype_7$predictedprob <- round(predict(LogitModel, newdata = prototype_7, type = "response")*100, digits = 1)

prototype_8$predictedprob <- round(predict(LogitModel, newdata = prototype_8, type = "response")*100, digits = 1)

prototype_9$predictedprob <- round(predict(LogitModel, newdata = prototype_9, type = "response")*100, digits = 1)

prototype_10$predictedprob <- round(predict(LogitModel, newdata = prototype_10, type = "response")*100, digits = 1)

prototype_11$predictedprob <- round(predict(LogitModel, newdata = prototype_11, type = "response")*100, digits = 1)

prototype_12$predictedprob <- round(predict(LogitModel, newdata = prototype_12, type = "response")*100, digits = 1)

```

__(D-1) Predict probabilities for prototypical individuals__
```{r warning=FALSE}
rbind.data.frame(prototype_1, prototype_2, prototype_3, prototype_4, prototype_5, 
    prototype_6, prototype_7, prototype_8, prototype_9, prototype_10, prototype_11, 
    prototype_12)
```

We can see that White man with % of other obligations has `r round(prototype_1$predictedprob,digits = 2)`% of chances to be approved for a loan. At the same situation, black people have, for the same situation only `r round(prototype_3$predictedprob,digits = 2)`% of chances to be approved for a loan and Hispanics have only `r round(prototype_2$predictedprob,digits = 2)`% of chances to be approved for a loan.

__(E) Estimate Probit Model__
```{r warning=FALSE}
ProbitModel = glm(Approve ~ Other_Obligations + Credit_History + Loan_Percentage + Married + Black + Hispan, data = base, family = "binomial" (link = "probit"))

summary(ProbitModel)
```

__The signals of the model suggest that:__

  1.	Increases in Other Obligation diminish the likelihood of approval of the loan, which makes sense.  
  2.	Being Black or Hispanic diminish your likelihood of approval of the loan, which does not make sense (your ethnicity does not interfere to your capacity to pay a loan), but was expected. 
  
__(F) Predict probabilities for prototypical individuals__
```{r warning=FALSE}

prototype_1$predictedprob <- round(predict(ProbitModel, newdata = prototype_1, type = "response")*100, digits = 1)

prototype_2$predictedprob <- round(predict(ProbitModel, newdata = prototype_2, type = "response")*100, digits = 1)

prototype_3$predictedprob <- round(predict(ProbitModel, newdata = prototype_3, type = "response")*100, digits = 1)

prototype_4$predictedprob <- round(predict(ProbitModel, newdata = prototype_4, type = "response")*100, digits = 1)

prototype_5$predictedprob <- round(predict(ProbitModel, newdata = prototype_5, type = "response")*100, digits = 1)

prototype_6$predictedprob <- round(predict(ProbitModel, newdata = prototype_6, type = "response")*100, digits = 1)

prototype_7$predictedprob <- round(predict(ProbitModel, newdata = prototype_7, type = "response")*100, digits = 1)

prototype_8$predictedprob <- round(predict(ProbitModel, newdata = prototype_8, type = "response")*100, digits = 1)

prototype_9$predictedprob <- round(predict(ProbitModel, newdata = prototype_9, type = "response")*100, digits = 1)

prototype_10$predictedprob <- round(predict(ProbitModel, newdata = prototype_10, type = "response")*100, digits = 1)

prototype_11$predictedprob <- round(predict(ProbitModel, newdata = prototype_11, type = "response")*100, digits = 1)

prototype_12$predictedprob <- round(predict(ProbitModel, newdata = prototype_12, type = "response")*100, digits = 1)
```

__(F) Predict probabilities for prototypical individuals__
```{r warning=FALSE}
rbind.data.frame(prototype_1, prototype_2, prototype_3, prototype_4, prototype_5, 
    prototype_6, prototype_7, prototype_8, prototype_9, prototype_10, prototype_11, 
    prototype_12)
```

We can see that White man with `r round(prototype_1$Other_Obligations,digits = 2)`% of other obligations has `r round(prototype_1$predictedprob,digits = 2)`% of chances to be approved for a loan. At the same situation, black people have, for the same situation only `r round(prototype_3$predictedprob,digits = 2)`% of chances to be approved for a loan and Hispanics have only `r round(prototype_2$predictedprob,digits = 2)`% of chances to be approved for a loan.  
The results are consistent between both models.

## 5.2 Model considering Gender

__(A) Estimate Logit Model__
```{r warning=FALSE}
LogitModel = glm(Approve ~ Other_Obligations + Male, data = base, 
                 family = "binomial")
summary(LogitModel)
```

__The signals of the model suggest that:__  

  1.	Increases in Other Obligation diminish the likelihood of approval of the loan, which makes sense.  
  2.	Being woman diminish your likelihood of approval of the loan, which does not make sense (your gender does not interfere to your capacity to pay a loan), but was expected. 
  
__(B) Generate Odds Ratios__
```{r warning=FALSE}
exp(coef(LogitModel))
```

__The Odd ratios suggest that:__

1.	Man had `r round(exp(coef(LogitModel))[3],digits=2)` times the odds of being approved for a loan as a woman;(p<0.01)   
2.	One unit of increase in other obligations reduce in `r round((1-exp(coef(LogitModel))[2])*100,digits=2)` percentage the odds of being approved for a loan in ;(p<0.01)    

__(C) Define prototypical loan applicants__
```{r warning=FALSE}
prototype_woman <- data.frame(Other_Obligations=mean(base$Other_Obligations),Male = 0)
prototype_woman <-prototype_woman %>% mutate(Male = as.factor(Male))
# levels(prototype_woman$Male) <- "Female"
prototype_men <- data.frame(Other_Obligations=mean(base$Other_Obligations),Male = 1)
prototype_men <- prototype_men %>% mutate(Male = as.factor(Male))
# levels(prototype_men$Male) <- "Male"
#Predict probabilities for prototypical individuals
prototype_woman$predictedprob <- round(
  predict (LogitModel,
           newdata = prototype_woman,
           type ="response")*100,
  digits = 1)


prototype_men$predictedprob <-  round(
  predict (LogitModel,
           newdata = prototype_men,
           type ="response")*100,
  digits = 1)
```

__(C-1) Define prototypical loan applicants - Female__
```{r warning=FALSE}
prototype_woman
```

__(C-2) Define prototypical loan applicants - Male__
```{r warning=FALSE}
prototype_men
```
Both Man and Woman have the similar probabilities of being approved for a loan considering they have `r round(prototype_men$Other_Obligations,digits = 2)`% of other obligation


__(D) Estimate Probit Model__
```{r warning=FALSE}
ProbitModel = glm(Approve ~ Other_Obligations + Male , data = base, 
                  family = "binomial" (link = "probit"))
summary(ProbitModel)
```

__The signals of the model suggest that:__
  1.	Increases in Other Obligation diminish the likelihood of approval of the loan, which makes sense.  
  2.	Being woman diminish your likelihood of approval of the loan, which does not make sense (your gender does not interfere to your capacity to pay a loan), but was expected. 
  
__(E) Predict probabilities for prototypical individuals__
```{r warning=FALSE}
prototype_woman$predictedprob <- round(
  predict (ProbitModel,
           newdata = prototype_woman,
           type ="response")*100,
  digits = 1)
prototype_men$predictedprob <-  round(
  predict (ProbitModel,
           newdata = prototype_men,
           type ="response")*100,
  digits = 1)
```

__(E-1) Predict probabilities for prototypical individuals - Female__
```{r warning=FALSE}
prototype_woman
```

__(E-2) Predict probabilities for prototypical individuals - Male__
```{r warning=FALSE}
prototype_men
```

The results are consistent between both models.